---
title: "R Notebook"
output: html_notebook
---

#### Setting up workspace 
```{r}
# supress warnings!!!

# Loading libraries
library(dplyr) # data manipulation
library(tidyr) # tidy data, e.g. pivot data
library(readxl) # read excel files
library(purrr)     # apply functions over lists
library(SHARK4R)   # match_worms_taxa — WoRMS taxonomy resolution
library(tidyverse) # 
library(stringr)   # str_trim, str_replace_all — string cleaning
library(vegan)
library(sf)        # For handling vector data
library(terra)     # For handling raster data
library(ggplot2)   # For data visualization
library(mapview)   # For interactive mapping
library(tmap) 
library(iNEXT) 
library(broom)
library(FactoMineR)
library(factoextra)
library(lmerTest) 

```
### Data
#### Reading in data
** function that reads in first survey data
```{r,message=FALSE}
# Survey Data
df_surveydata <- read_speciesdata()

    # Turn year into numeric for further analysis 
    df_surveydata <- df_surveydata %>%
      mutate(year = as.numeric(year))

    # Cleaned table for reference 
    
# Grab site information
df_grabdata <- read_grabdata() #records = df_surveydata)
```

#### Presence and abscence 
```{r}
# create new dataframe (just 1 and 0)
df_P_A_surveydata <- function_presence_and_absence(survey_data = df_surveydata)

df_meta_PA <- prepare_community_data(df_survey = df_P_A_surveydata,
                                  df_env = df_grabdata,
                                  exclude_samples = c("G28_2013", "T2_01_2022"))

combined_table_PA <- df_meta_PA$combined
```

### Data Prep for Analysis 
```{r}
# Preparing data for analysis- creating a combined species data and grab data
  # only species included
df_meta <- prepare_community_data(df_survey = df_surveydata,
                                  df_env = df_grabdata,
                                  exclude_samples = c("G28_2013", "T2_01_2022"))

combined_table <- df_meta$combined
```

### BTA (Objective 2)
#### BTA - Making Template
```{r}
# Making the BTA template files that have species matched with biotic 
BTA_template <- build_empty_trait_matrix(survey_data = df_surveydata, 
                                         output_dir = "/Users/sarahstevenson/Desktop/diss/SouthArran/outputs/abundance",
                                         biotic_path= "data/biotic.csv")

# Extracting trait modalities 
trait_groups   <- BTA_template$trait_groups
all_trait_cols <- unlist(trait_groups, use.names = FALSE)

# Making the BTA template files that have species matched with biotic (PRESENCE and ABSENCE)
BTA_P_A_template <- build_empty_trait_matrix(survey_data = df_P_A_surveydata,
                                             output_dir = "/Users/sarahstevenson/Desktop/diss/SouthArran/outputs/P_A",
                                             biotic_path= "data/biotic.csv")
```

#### BTA - Reading in filled trait matrix
```{r}
# Reading in the filled trait by taxon matrix 
filled_matrix <- read_csv(
  "/Users/sarahstevenson/Desktop/diss/SouthArran/outputs/abundance/trait_matrix_FILLED.csv",
  col_types = cols(
    .default = col_guess(),
    !!!setNames(rep(list(col_double()), length(all_trait_cols)), all_trait_cols)))

# Verifying BTA matrix is filled in correctly
cleaned_matrix <- check_trait_matrix(filled_matrix, trait_groups = trait_groups)

```

#### BTA - Analysis
##### Temporal 
```{r}
# With abundance (species count)
  # Need names to match 
    df_surveydata <- df_surveydata %>%
      mutate(bta_name = case_when(
        !is.na(accepted_name) & accepted_name != "" ~ accepted_name,
        TRUE ~ SpeciesName_clean))

# Running PCA
results_PCA_abundunce <- run_bta_pca_workflow(
  filled_matrix  = filled_matrix,
  trait_groups   = trait_groups,
  abundance_data = df_surveydata,
  metadata_data  = df_meta$meta,
  top_n_contrib  = 10,
  taxon_col      = "bta_name",
  sample_col     = "sample_id",
  period_col     = "period",
  phi_col        = "Mean_phi",
  depth_col      = "Depth",
  colors = c("Before" = "#1C77C3", "After" = "#FB8500"))


# Sediment environment focus (Survey stations T)
result_T <- run_bta_pca_workflow(
  filled_matrix   = filled_matrix,        
  trait_groups    = trait_groups,
  abundance_data  = df_surveydata,       
  metadata_data   = df_meta$meta,           
  taxon_col       = "bta_name",
  sample_col      = "GrabSite",
  abund_col       = "value",
  station_col     = "GrabSite_station",
  station_filter  = "T",        # Selects stations starting with "T"
  period_col      = "period",
  phi_col         = "Mean_phi",
  depth_col       = "Depth",
  top_n_contrib   = 10,
  colors = c("Before" = "#1C77C3", "After" = "#FB8500"))


# Presence and absence (0 and 1 only)


```

### Community Structure Analysis (Objective 1)
##### Temporal 
```{r}
#Obj 1 part 1
results_nmds <- run_nmds_analysis(df_meta,
                              group_col = "period",
                              covariates = c("Mean_phi", "Depth"),
                              run_simper = TRUE,
                              colors = c("Before" = "#1C77C3", "After" = "#FB8500"))

print(results_nmds$plot)
results_nmds$permanova
results_nmds$betadisper_anova
results_nmds$simper

# Obj1 part 2
diversity_measurements <- analyses_bio_patterns(
  comm        = df_meta$matrix,
  meta        = df_meta$meta,
  period_col  = "period",
  covariates  = c("Mean_phi", "Depth"),
  station_col = NULL) # moving forward can account for survey stations but to retain 2012/13 data it is set to NULL

diversity_measurements$summary # coefficient table for all 3 metrics

# species_richness <- diverstity_measurements$summary %>%
  # dplyr::filter(response == "Richness") 
    #to get just species richness and save as an object

# Rounding values for inspection 
diversity_measurements$summary %>%
  mutate(
    estimate  = signif(estimate, 3),
    std.error = signif(std.error, 3),
    statistic = round(statistic, 2),
    p.value   = case_when(
      p.value < 0.001 ~ "< 0.001",
      TRUE ~ as.character(round(p.value, 3))))

```

### Spatial Analysis (Objective 3)
##### Adding in protection zones, changing sediment type and depth into categorical variables
```{r}
# Reading in protection levels
mpa_restrictions <- st_read("data/Fishing measures relevant to South Arran/Fishing measures relevant to South Arran.shp")

if (!all(c("protection_level", "depth_category", "sediment_type_simple") %in% names(df_grabdata))) { # adding in safety block
  
# Adding all levels to environmental dataframe
  df_grabdata <- add_zones_protect_depth_sedi(
    df_env        = df_grabdata,
    mpa_shapefile = mpa_restrictions,
    df_sites      = df_grabdata,
    site_col      = "GrabSite",
    lon_col       = "Longitude",
    lat_col       = "Latitude")
}

spatial_result <- prepare_community_data(df_surveydata, df_grabdata)
spatial_result$meta %>% select(sample_id, GrabSite, protection_level, period, depth_category, sediment_type_simple ) 
```

### Community Structure Analysis (Objective 1)
##### Spatial
```{r}
## NOTE: period_col here in place of management period is protection levels 

# Obj 1 part 1 - by protection level
results_nmds_protection <- run_nmds_analysis(spatial_result,
                              group_col = "protection_level",
                              covariates = c("Mean_phi", "Depth"),
                              exclude_samples = c("T2_01_2022", "G28_2013"),
                              run_simper = TRUE,
                              colors = c("High" = "#0000CC",
                                                   "Medium" = "#3366FF",
                                                   "Low" = "#99FFFF",
                                                   "Outside MPA"= "#CCCCCC"))

print(results_nmds_protection$plot)
results_nmds_protection$permanova
results_nmds_protection$betadisper_anova
results_nmds_protection$simper_top10_overall

exclude_ids <- c("T2_01_2022", "G28_2013")
keep <- !(spatial_result$meta$sample_id %in% exclude_ids)

# Biodiversity measures
diversity_measurements_protection <- analyses_bio_patterns(
  comm        = spatial_result$matrix[keep, , drop = FALSE],
  meta        = spatial_result$meta[keep, , drop = FALSE],
  period_col  = "protection_level",
  covariates  = c("Mean_phi", "Depth"),
  station_col = NULL) 

diversity_measurements_protection$summary

# For viewing
diversity_measurements_protection$summary %>%
  mutate(
    estimate  = signif(estimate, 3),
    std.error = signif(std.error, 3),
    statistic = round(statistic, 2),
    p.value   = case_when(
      p.value < 0.001 ~ "< 0.001",
      TRUE ~ as.character(round(p.value, 3))))
```

#### BTA - Analysis
##### Spatial 
```{r}
## NOTE: period_col here in place of management period is protection levels 

results_PCA_abundance_protection <- run_bta_pca_workflow(
  filled_matrix  = filled_matrix,
  trait_groups   = trait_groups,
  abundance_data = df_surveydata,
  metadata_data  = spatial_result$meta,
  top_n_contrib  = 10,
  taxon_col      = "bta_name",
  sample_col     = "sample_id",
  period_col     = "protection_level",  # passing protection_level here instead of period
  phi_col        = "Mean_phi",
  depth_col      = "Depth",
  colors = c("High" = "#0000CC",
             "Medium" = "#3366FF",
              "Low" = "#99FFFF",
              "Outside MPA"= "#CCCCCC"))
```

### Species Accumulation Curve (Survey Effort Analysis)
```{r}
# Cleaning presence and absence data 
df_meta_PA <- prepare_PA_data(df_survey = df_P_A_surveydata,
                              df_env = df_grabdata,
                              exclude_samples = c("G28_2013", "T2_01_2022"))

PA_combined_table <- df_meta_PA$combined

# Species Accumulation Curve 
sac_output <- run_station_sac(PA_combined_table)
sac_output$plot

sac_output$inext_result


```
